Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-02-03 Multiresolution_framework/ltx/manuscript.tex
minor edit
minor edit to the appendix

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}),
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r},
\end{align}
to give the state transition equation
\begin{equation}\label{eq:StateEquation}
 \mathbf x_{t+1} =\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t,
\end{equation} 
where
\begin{equation}\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta)= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{equation} 
The problem of joint state and parameter estimation for the state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation} can be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations parameterized  by $\boldsymbol\theta$, known as the likelihood function. This problem cannot be solved separately as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known Expectation Maximization (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. There are two main applications of the EM algorithm:  first is when the observations are incomplete due to a faulty observation process or existing limitations and the second  when the direct maximization of the likelihood function is difficult but can be simplified by assuming the existence of an additional however missing (or hidden) variables \cite{Bilmes1998}, the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \cite{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

@@

 +\sum_{t=0}^{T-1}2\ln p(\mathbf x_0)\right],

@@

is constant with respect to parameter $\boldsymbol\theta$ and will disappear subject to differentiation with respect to $\boldsymbol\theta$. Taking the trace and rearranging, using the invariant cyclic permutations property of the trace, this distribution can be written as

@@

The EM algorithm alternates between maximizing the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\theta'$, in the E-step the lower bound is maximized with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximizing the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\theta,\boldsymbol\theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \theta'}\left[ \cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximizes the expectation of the complete data log-likelihood. 
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}\label{ap:Xiderivation}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in equation~\eqref{eq:Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	\end{equation}                                                                                                
	We start by expanding the equation \eqref{eq:appXi0}, giving

>>>>

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}),
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r},
\end{align}
to give the state equation
\begin{equation}\label{eq:StateEquation}
 \mathbf x_{t+1} =\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t,
\end{equation} 
with the parameterized transition matrix
\begin{equation}\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta)= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{equation} 
The problem of joint state and parameter estimation for the state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation} can be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations parameterized  by $\boldsymbol\theta$, known as the likelihood function. This problem cannot be solved separately as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known Expectation Maximization (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. There are two main applications of the EM algorithm:  first is when the observations are incomplete due to a faulty observation process or existing limitations and the second  when the direct maximization of the likelihood function is difficult but can be simplified by assuming the existence of an additional however missing (or hidden) variables \cite{Bilmes1998}, the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \cite{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

@@

 +\sum_{t=0}^{T-1}2\ln p(\mathbf x_0)\right].

@@

is constant with respect to the parameter $\boldsymbol\theta$, and will disappear subject to differentiation with respect to $\boldsymbol\theta$. Taking the trace and rearranging, using the invariant cyclic permutations property of the trace, this distribution can be written as

@@

The EM algorithm alternates between maximizing the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\theta'$, in the E-step the lower bound is maximized with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximizing the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\theta,\boldsymbol\theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \theta'}\left[ \cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximizes the expectation of the complete data log-likelihood. 
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}\label{ap:Xiderivation}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in equation~\eqref{eq:MRA-Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	\end{equation}                                                                                                
	We start by expanding the equation \eqref{eq:appXi0}, giving
